### 2017-12-07 paper/sexSelectionSpacing-ver1.tex
Add method requirements to beginning of Estimation Strategy

Although it might be tempting to simply examine how observed
spacing has changed over time this is problematic.

Possible ways of doing this:
- Only closed spells (where a birth is observed)
    - can be split by the outcome (ending birth is either male or female)
    - can be split by the sex composition of prior children
- All spells (whether closed or not)
    - can only be split by the sex composition of prior children

Problems:
- Using closed spells only ignore that those families with the strongest
  son preference are also likely to have the longest spacing and therefore be 
  less likely to show up as a complete spell.
- Because of the potential for changes in the use of sex selection within a spell
  it is possible that parents go through multiple sex-selective abortions and still
  end up with a girl
- Some open spells may not result in a birth (childbearing finished with prior birth),
  which is especially a problem for higher parities where sex composition of
  prior children has a large effect on the probability of parity progression.

Need a method that accounts for censoring.

In addition to accounting for censoring, the empirical method must be able 
to capture the three main implications of the availability of sex-selective 
abortions combined with son preference:
a higher probability that the next child is a son, 
longer spacing to next birth with use of sex selection, 
and that the use of prenatal sex determination may change between 
one observed birth and the next.
In addition, it should be able to capture the shorter spacing between
births observed in the absence of use or availability of prenatal 
sex determination.
This means an empirical model that can jointly estimate 
the association between covariates and both the sex of children born 
and the spacing from last birth. 
My proposed model is a discrete time, non-proportional competing risk 
hazard model with two exit states: either a boy or a girl is born.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

I divide a woman's reproductive life into spells that each covers 
the period between births (or from marriage to first birth for the first spell).
For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$ and 
the spell continues until time $t_i$, when either a birth occurs or the spell 
is censored.%
\footnote{
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
}
There are two exit states: birth of a boy, $j=1$, or birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete time hazard rate $h_{ijt}$ is
\begin{equation}
 h_{ijt} = \Pr (T_i=t, J_i=j \mid T_i \geq t; \mathbf{Z}_{it},\mathbf{X}_{i} ),
\end{equation}
where $T_i$ is a discrete random variable that captures when woman $i$'s birth occurs.
To ease presentation the indicator for spell number is suppressed.
The explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, include 
individual, household, and community characteristics discussed below.

@@

The standard proportionality of hazards assumption does not work here.
First, even in the absense of sex selection, it is unlikely that the shape of
hazard functions are identical across groups.
[sex composition]

[TK I am not very satisfied with this explanation of the problems with
proportionality and the associated bias; non-proportionality is required
to capture changes in use of sex selection]

In principle, specifying the model as a proportional hazard model, i.e.\ one 
where covariates simply shift the hazard rates up or down independent of 
spell length, is more efficient, but only provided that the proportionality 
assumption holds.
If the proportionality assumption does not hold, however, the result is
a potentially substantial bias in estimates.
The problem is that a proportional hazard model does not allow covariates
to have different effects at different times within a spell and therefore
cannot capture differences in the shape of the hazard functions between 
different groups. 
It is highly unlikely---even in the absence of prenatal sex determination---that 
the baseline hazards are the same across education levels, areas of residence, 
or sex composition of previous births.
Any bias from the proportionality assumption is likely exacerbated by the 
introduction of prenatal sex determination for two reasons.
First, one of the main points of this paper is that use of sex-selective 
abortions affects birth spacing, and use of sex selection differs across groups.
Second, as discussed above, the use of sex selection may vary within a spell,
depending on the length of the spell.

I therefore use a non-proportional model where the main explanatory variables 
and the interactions between them are interacted with the baseline hazards.
This is captured by the $Z$ set of explanatory variables
\begin{equation}
 \mathbf{Z}_{it} = D_j(t) \times (\mathbf{Z}_1 + Z_2 + \mathbf{Z}_1 \times Z_2),
\end{equation}
where $D_j(t)$ is the piece-wise linear baseline hazard and $\mathbf{Z}_1$ captures sex 
composition of previous children, if any, and $Z_2$ captures area of residence.
This allows the effects of the main explanatory variables on the probabilities 
of having a boy, a girl, or no birth to vary over time within a spell.
The use of a non-proportional specification, together with a flexible baseline hazard, 
also mitigates any potential effects of unobserved heterogeneity \citep{Dolton1995}.

The remaining explanatory variables, $\mathbf{X}$, enter proportionally.
To further minimize any potential bias from assuming proportionality, estimations 
are done separately for different levels of mothers' education and for different 
time periods.
The exact specifications and the individual variables are described below.

Equation (\ref{eq:hazard}) is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, if the data are transformed so the unit of analysis is spell unit rather 
than the individual woman, the model can be estimated using a standard multinomial 
logit model.%
\footnote{
A potentially issue is that the multinomial model assumes that alternative 
exit states are stochastically independent,
also known as the Independence of Irrelevant Alternatives (IIA) assumption.
This assumption rules out any individual-specific unmeasured or 
unobservable factors that affect both the hazard of having a girl and the 
hazard of having a boy.
To address this issue the estimations include a proxy for fecundity
discussed in Section \ref{sec:data}.
In addition, the multivariate probit model can be used as an alternative
to the multinomial logit because the IIA is not imposed \citep{han90}.
The results are essentially identical between these two models and
available upon request.
}
In the reorganized data the outcome variable is zero if the
woman does not have a child in a given period, one if she gives birth to a son 
in that period, and two if she gives birth to a daughter in that period.

The main downside of this approach is that direct interpretation of the estimated 
coefficients for this model is challenging because of the competing risk setup.
First, coefficients show the change in hazards relative to the base outcome, 
no birth, rather than simply the hazard of an event.
Second, a positive coefficient does not necessarily imply that an increase in a
variable's value increases the probability of the associated event because the 
probability of another event may increase even more \citep{thomas96}.

@@

The other important part of the model is the spacing between births.
Spacing is captured by the survival curve, which shows the probability of not 
having had a birth yet by spell duration.
The survival curve at time $t$ is 
\begin{equation}
S_{t} 
= 
\prod_{d=1}^t 
\left( 
	1- \left(P(b_{d} | \mathbf{X}_{k}, \mathbf{Z}_{kd}, d) 
	+ P(g_{d} | \mathbf{X}_{k}, \mathbf{Z}_{kd}, d) \right) 
\right),
 \label{eq:survival}
\end{equation}
or equivalently
\begin{equation}
S_{t} 
= 
\prod_{d=1}^t
\left(
\frac{ 1 }
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{ld}'\mathbf{Z}_{kd} + \beta_l'\mathbf{X}_{k})}
\right).
\end{equation}

In the absence of sex selection, I expect most parities to show an 
``inverted s'' pattern, where there initially are relatively few births, 
followed by a substantial number of births over a 1 to 2 year period, and then
relatively few births thereafter.
As sex selection becomes more widely used the associated
longer spacing shows up by making the survival curve straighter,
indicating that some of the births that would originally have taken place
now take place later because of abortions.
The use of sex selection is clearly not the only factor that can change the shape of
the survival curves; factors such as the desired number of children and
use of contraceptives may also shape the shape.
To account for this it is best to compare survival curves for an individual
parity between women who are likely to use sex selection and women who are
not, for example because they already have one or more sons.

In addition to information about spacing, the survival curves also provide 
``weighting'' for the associated percentage boys born.
The steeper the survival curve, the more weight should be assigned to a given
spell period because it is based on more births, 
whereas a period with a flat survival curve should be given little weight because the 
percentage boys is based on few births.
Hence, although they cannot by themselves show sex selection, the survival
curves are a crucial complement to the estimated sex ratios by duration.

@@

>>>>

Requirements for method and the intuition behind each:  
 - Account for censoring (hazard model)
 - Work both with and without sex selection (competing risk part)
 - Allow for changes in use of sex selection within spell (flexible specification of baseline hazard ?)
 - Capture that the shape of the hazard function differ across groups
   (non-proportionality); needs a better description, maybe using example
   of shorter spacing and/or use of sex selection. Differences in parity
   progression likelihood.

I divide a woman's reproductive life into one or more spells.
Each spell covers the period from one birth, or marriage for the first spell,
to the following birth; I define how the duration is measured in more 
detail in Section \ref{sec:data}.
A basic problem is that some spells' durations are censored, either
because the survey took place before the end birth for the spell or because
the couple has finished childbearing.
The standard approach, which I follow here, is to use a hazard models to 
address the censoring of spells.

Prior research on birth spacing have used hazard models where the exit 
state was the birth of a child. 
This works when there is no sex selection---making the sex of the child 
a random event---but not here because at least some couples use 
sex selection. 
To capture that the sex of children are no longer a random event, I instead
use a competing risk framework with two possible exit states: either a boy 
or a girl is born.%
\footnote{
\cite{Merli2000} used a discrete hazard model to examine whether 
there were under-reporting of births in rural China, although they 
estimated separate waiting time regressions for boys and girls.
}

Proportional hazard models, where covariate have an multiplicative 
effect on the hazard rate, are more efficient than non-proportional 
models, but only provided that the proportionality assumption holds.
The problem here is that a proportional hazard model does not allow covariates
to have different sized effects at different times within a spell and therefore
cannot capture differences in the shape of the hazard functions between 
different groups, and if the proportionality assumption does not hold 
the result is a potentially substantial bias in estimates.
It is unlikely---even in the absence of prenatal sex 
determination---that the effect of, for example, the sex composition
of prior births have the same effect throughout the entire spell.
[son preference and parity progression also leads to differences!]
Furthermore, any bias from the proportionality assumption is likely 
exacerbated by the introduction of prenatal sex determination.
First, sex-selective abortions affects birth spacing and its use
vary across groups.
Second, the use of sex selection may vary within a spell, which means 
that the effect of covariate may vary within the spell as well.
Finally, the use of a non-proportional specification, together with a 
flexible baseline hazard, also mitigates any potential effects of 
unobserved heterogeneity \citep{Dolton1995}.
My proposed model is a discrete time, non-proportional competing risk 
hazard model with two exit states: either a boy or a girl is born.

For each woman, $i=1,\ldots,n$, the starting point for a spell is time $t=1$ and 
the spell continues until time $t_i$, when either a birth occurs or the spell 
is censored.%
\footnote{
The time of censoring is assumed independent of the hazard rate,
as is standard in the literature.
}
There are two exit states: birth of a boy, $j=1$, or birth of a girl, $j=2$, and 
$J_i$ is a random variable indicating which event took place.
The discrete time hazard rate $h_{ijt}$ is
\begin{equation}
 h_{ijt} = \Pr (T_i=t, J_i=j \mid T_i \geq t; \mathbf{Z}_{it},\mathbf{X}_{i} ),
\end{equation}
where $T_i$ is a discrete random variable that captures when woman $i$'s birth occurs.
To ease presentation the indicator for spell number is suppressed.
The explanatory variable vectors, $\mathbf{Z}_{it}$ and $\mathbf{X}_{i}$, include 
individual, household, and community characteristics discussed below.

@@

The $Z$ set of explanatory variable and the interactions between them 
are the non-proportional part of the model, which means that they are
interacted with the baseline hazard:
\begin{equation}
 \mathbf{Z}_{it} = D_j(t) \times (\mathbf{Z}_1 + Z_2 + \mathbf{Z}_1 \times Z_2),
\end{equation}
where $D_j(t)$ is the piece-wise linear baseline hazard and $\mathbf{Z}_1$ captures sex 
composition of previous children, if any, and $Z_2$ captures area of residence.
This allows the effects of the main explanatory variables on the probabilities 
of having a boy, a girl, or no birth to vary over time within a spell.
To further minimize any potential bias from assuming proportionality, estimations 
are done separately for different levels of mothers' education and for different 
time periods.
The remaining explanatory variables, $\mathbf{X}$, enter proportionally.
% The exact specifications and the individual variables are described below.

Equation (\ref{eq:hazard}) is equivalent to the logistic hazard model and has the same 
likelihood function as the multinomial logit model \citep{allison82,jenkins95}.
Hence, if the data are transformed so the unit of analysis is spell unit, in 
this case 3 months periods, rather than the individual woman, the model can be 
estimated using a standard multinomial logit model.%
\footnote{
A potentially issue is that the multinomial model assumes that alternative 
exit states are stochastically independent,
also known as the Independence of Irrelevant Alternatives (IIA) assumption.
This assumption rules out any individual-specific unmeasured or 
unobservable factors that affect both the hazard of having a girl and the 
hazard of having a boy.
To address this issue the estimations include a proxy for fecundity
discussed in Section \ref{sec:data}.
In addition, the multivariate probit model can be used as an alternative
to the multinomial logit because the IIA is not imposed \citep{han90}.
The results are essentially identical between these two models and
available upon request.
}
In the reorganized data the outcome variable is zero if the
woman does not have a child in a given period, one if she gives birth to a son 
in that period, and two if she gives birth to a daughter in that period.

The main downside is that direct interpretation of the estimated 
coefficients is challenging because of the competing risk setup.
First, coefficients show the change in hazards relative to the base outcome, 
no birth, rather than simply the hazard of an event.
Second, a, for example, positive coefficient does not necessarily imply that
a higher variable value is associated with a higher event probability because the 
probability of another event may increase even more \citep{thomas96}.

@@

The distribution of spacing is captured by the survival curve, which shows 
the probability of not having had a birth yet by spell duration.
The survival curve at time $t$ is 
\begin{equation}
S_{t} 
= 
\prod_{d=1}^t 
\left( 
	1- \left(P(b_{d} | \mathbf{X}_{k}, \mathbf{Z}_{kd}, d) 
	+ P(g_{d} | \mathbf{X}_{k}, \mathbf{Z}_{kd}, d) \right) 
\right),
 \label{eq:survival}
\end{equation}
or equivalently
\begin{equation}
S_{t} 
= 
\prod_{d=1}^t
\left(
\frac{ 1 }
{1 + \sum_{l=2}^2 \exp(D_j(t) + \alpha_{ld}'\mathbf{Z}_{kd} + \beta_l'\mathbf{X}_{k})}
\right).
\end{equation}

@@

[Moved from estimation strategy]
In the absence of sex selection, I expect most parities to follow
a pattern, where there initially are relatively few births, 
followed by a substantial number of births over a 1 to 2 year period, and then
relatively few births thereafter.
As sex selection becomes more widely used the associated
longer spacing shows up by making the survival curve straighter,
indicating that some of the births that would originally have taken place
now take place later because of abortions.
The use of sex selection is clearly not the only factor that can change the shape of
the survival curves; factors such as the desired number of children and
use of contraceptives may also shape the shape.
To account for this it is best to compare survival curves for an individual
parity between women who are likely to use sex selection and women who are
not, for example because they already have one or more sons.

[Moved from estimation strategy]
In addition to information about spacing, the survival curves also provide 
``weighting'' for the associated percentage boys born.
The steeper the survival curve, the more weight should be assigned to a given
spell period because it is based on more births, 
whereas a period with a flat survival curve should be given little weight because the 
percentage boys is based on few births.
